---
title: "Session 1: Relational thinking"
subtitle: "Core concepts in network science"
date: today
---

## Goals

By the end of this session you can

- explain the difference between **attribute data** and **relational data**,
- name the building blocks of a network: nodes, edges, direction, weight,
- say why standard regression assumptions break down when observations are connected.

## 1. Think first: attributes vs. relations

::: {.callout-note title="Think first (pen and paper, 10 min)"}
Take five countries you know well. Write down

1. one **attribute** of each country (e.g. GDP, regime type), and
2. one **relation** between pairs of countries (e.g. "shares a border with").

How many values did you need for (1)? How many *possible* values for (2)? What happens to that number with 190 countries?
:::

```{r}
library(tidyverse)
library(igraph)
```

### Attribute data: one row per country

This is the data format you know from regression: each row is a unit, each column a property of that unit.

```{r}
countries <- tribble(
  ~country, ~eu_member, ~landlocked,
  "AUT",    TRUE,       TRUE,
  "CHE",    FALSE,      TRUE,
  "DEU",    TRUE,       FALSE,
  "FRA",    TRUE,       FALSE,
  "ITA",    TRUE,       FALSE
)
countries
```

### Relational data: one row per pair

Now the relation "shares a border". The unit of observation is no longer a country but a **pair of countries** (a dyad). With five countries there are 10 possible pairs:

```{r}
dyads <- combn(countries$country, 2) |>
  t() |>
  as_tibble(.name_repair = ~ c("country_1", "country_2"))

borders <- tribble(
  ~country_1, ~country_2,
  "AUT", "CHE",
  "AUT", "DEU",
  "AUT", "ITA",
  "CHE", "DEU",
  "CHE", "FRA",
  "CHE", "ITA",
  "DEU", "FRA",
  "FRA", "ITA"
)

dyads <- dyads |>
  left_join(borders |> mutate(border = 1), by = c("country_1", "country_2")) |>
  mutate(border = replace_na(border, 0))
dyads
```

::: {.callout-note title="Think first"}
Which pairs have `border = 0`? Notice that "no border" is information too. An edge list (Section 3) usually keeps only the 1s, so the 0s are implicit.
:::

### How fast do pairs grow?

```{r}
n <- c(5, 10, 50, 190)
tibble(
  countries = n,
  undirected_pairs = n * (n - 1) / 2,
  directed_pairs = n * (n - 1)
)
```

With 190 countries you have 190 rows of attribute data, but almost 18,000 possible undirected pairs.

### Putting both together: a network

A network combines the two. Relations become **edges**, and attributes stay attached to the **nodes**.

```{r}
g_borders <- graph_from_data_frame(borders, directed = FALSE, vertices = countries)

plot(
  g_borders,
  vertex.color = ifelse(V(g_borders)$eu_member, "steelblue", "grey80"),
  vertex.label.color = "black"
)
```

::: {.callout-note title="Think first"}
Switzerland is the only non-EU country here, yet it borders all four others. What does that tell you about using attributes alone to understand a country's position in Europe?
:::

## 2. Building blocks

A **network** (or graph) consists of

- **nodes** (vertices): the units, e.g. states, organisations, leaders,
- **edges** (ties, links): the relations between them.

Edges can be

- **undirected** (A–B is the same as B–A: "share a border") or **directed** (A→B is not B→A: "sends aid to"),
- **binary** (present or absent) or **weighted** (e.g. trade volume).

## 3. Your first network

The simplest way to write down a network is to list who is connected to whom.

```{r}
edges <- tribble(
  ~from, ~to,
  "A", "B",
  "A", "C",
  "B", "C",
  "C", "D",
  "D", "E"
)

g <- graph_from_data_frame(edges, directed = FALSE)
plot(g)
```

::: {.callout-note title="Think first"}
Before running the next chunk: how many ties does each node have? Which node would you call most "central", and why?
:::

```{r}
degree(g)
```

Now compare: is the node with the most ties also the one whose removal would split the network?

```{r}
plot(delete_vertices(g, "C"))
plot(delete_vertices(g, "D"))
```

Already here, "centrality" means more than one thing. We return to this in Session 4.

## 4. Why relations break standard statistics

Regression assumes observations are **independent**. In a network, a tie between A and B is not independent of ties A–C and B–C: if two of your allies are allied with each other, that changes the picture. This **dependence** is not a nuisance in network science; it is the object of study. Sessions 5 and 6 show how to model it.

## Your turn

::: {.callout-tip title="Exercise"}
1. Build a directed network of five actors where *one* actor sends ties to everybody and nobody sends ties back. Plot it.
2. Compute `degree(g, mode = "in")` and `degree(g, mode = "out")`. Explain the result in one sentence each.
3. Name one IR phenomenon where direction matters and one where it does not.
:::

```{r}
# your code here
```
