---
title: "Session 1: Relational thinking"
subtitle: "Core concepts in network science"
date: 2027-01-07
---

## Goals

By the end of this session you can

- explain the difference between **attribute data** and **relational data**,
- name the building blocks of a network: nodes, edges, direction, weight,
- say why standard regression assumptions break down when observations are connected.

## 1. Think first: attributes vs. relations

::: {.callout-note title="Think first (pen and paper, 10 min)"}
Take five countries you know well. Write down

1. one **attribute** of each country (e.g. GDP, regime type), and
2. one **relation** between pairs of countries (e.g. "shares a border with").

How many values did you need for (1)? How many *possible* values for (2)? What happens to that number with 190 countries?
:::

Attribute data has one row per **unit**. Relational data has one row per **pair** of units, and the number of possible pairs grows much faster than the number of units.

```{r}
library(tidyverse)
library(igraph)

n <- c(5, 10, 50, 190)
tibble(
  countries = n,
  undirected_pairs = n * (n - 1) / 2,
  directed_pairs = n * (n - 1)
)
```

## 2. Building blocks

A **network** (or graph) consists of

- **nodes** (vertices): the units, e.g. states, organisations, leaders,
- **edges** (ties, links): the relations between them.

Edges can be

- **undirected** (A–B is the same as B–A: "share a border") or **directed** (A→B is not B→A: "sends aid to"),
- **binary** (present or absent) or **weighted** (e.g. trade volume).

## 3. Your first network

The simplest way to write down a network is to list who is connected to whom.

```{r}
edges <- tribble(
  ~from, ~to,
  "A", "B",
  "A", "C",
  "B", "C",
  "C", "D",
  "D", "E"
)

g <- graph_from_data_frame(edges, directed = FALSE)
plot(g)
```

::: {.callout-note title="Think first"}
Before running the next chunk: how many ties does each node have? Which node would you call most "central", and why?
:::

```{r}
degree(g)
```

Now compare: is the node with the most ties also the one whose removal would split the network?

```{r}
plot(delete_vertices(g, "C"))
plot(delete_vertices(g, "D"))
```

Already here, "centrality" means more than one thing. We return to this in Session 4.

## 4. Why relations break standard statistics

Regression assumes observations are **independent**. In a network, a tie between A and B is not independent of ties A–C and B–C: if two of your allies are allied with each other, that changes the picture. This **dependence** is not a nuisance in network science; it is the object of study. Sessions 5 and 6 show how to model it.

## Your turn

::: {.callout-tip title="Exercise"}
1. Build a directed network of five actors where *one* actor sends ties to everybody and nobody sends ties back. Plot it.
2. Compute `degree(g, mode = "in")` and `degree(g, mode = "out")`. Explain the result in one sentence each.
3. Name one IR phenomenon where direction matters and one where it does not.
:::

```{r}
# your code here
```
