---
title: "Session 3: Preparing and handling network data"
subtitle: "From messy raw data to a clean network object"
date: 2027-01-21
---

## Goals

- spot the typical problems in raw relational data (inconsistent IDs, duplicates, self-loops, missing endpoints, isolates),
- clean an edge list step by step and justify each decision,
- build network objects from separate node and edge lists, including isolates,
- slice a network over time,
- work with networks in the tidy style using **tidygraph**.

```{r}
library(tidyverse)
library(igraph)
library(tidygraph)
library(here)
```

## 1. Messy raw data

Real relational data rarely arrives clean. Here is a small example with the problems you will meet in practice:

```{r}
raw <- tribble(
  ~from,  ~to,   ~year,
  "AUT",  "DEU", 2010,
  "aut ", "DEU", 2010,   # inconsistent spelling of the same ID
  "DEU",  "AUT", 2010,   # same undirected tie, reversed
  "FRA",  "FRA", 2010,   # self-loop
  "ITA",  NA,    2011,   # missing endpoint
  "USA",  "GBR", 2011,
  "GBR",  "USA", 2012
)
raw
```

::: {.callout-note title="Think first"}
Before cleaning: how many *real* ties does this data contain? Which rows are problems, and is each one an error or a substantive decision?
:::

## 2. Cleaning step by step

```{r}
clean <- raw |>
  # 1. harmonise IDs
  mutate(across(c(from, to), \(x) toupper(str_trim(x)))) |>
  # 2. drop rows with a missing endpoint
  filter(!is.na(from), !is.na(to)) |>
  # 3. drop self-loops
  filter(from != to) |>
  # 4. undirected: put each pair in a fixed order, then remove duplicates
  mutate(a = pmin(from, to), b = pmax(from, to)) |>
  distinct(a, b, year) |>
  rename(from = a, to = b)

clean
```

::: {.callout-warning title="Discuss"}
Step 4 is only correct if the relation is undirected. What would you do if this were a directed relation, e.g. arms transfers? Is a self-loop always an error? (Think of a state's trade with itself in some datasets.)
:::

## 3. Node lists and isolates

An edge list cannot contain states with **no** ties. If isolates matter (they almost always do in IR), you need a separate node list:

```{r}
nodes <- tibble(
  name = c("AUT", "DEU", "FRA", "GBR", "USA", "ITA", "CHE"),
  eu_member = c(TRUE, TRUE, TRUE, FALSE, FALSE, TRUE, FALSE)
)

g_without <- graph_from_data_frame(clean, directed = FALSE)
g_with <- graph_from_data_frame(clean, directed = FALSE, vertices = nodes)

vcount(g_without)
vcount(g_with)
```

::: {.callout-note title="Think first"}
Density is the share of possible ties that exist. How does leaving out isolates change the density? Compute both by hand, then check with `edge_density()`.
:::

```{r}
edge_density(g_without)
edge_density(g_with)
```

## 4. Networks over time

Panel network data needs to be sliced: one network per year (or per period).

```{r}
networks_by_year <- clean |>
  group_split(year) |>
  map(\(e) graph_from_data_frame(e, directed = FALSE, vertices = nodes))

names(networks_by_year) <- sort(unique(clean$year))
map_dbl(networks_by_year, edge_density)
```

## 5. The tidy way: tidygraph

`tidygraph` lets you manipulate nodes and edges with dplyr verbs. `activate()` switches between the node table and the edge table.

```{r}
tg <- tbl_graph(nodes = nodes, edges = clean, directed = FALSE)
tg

tg |>
  activate(nodes) |>
  mutate(degree = centrality_degree()) |>
  arrange(desc(degree))
```

## 6. Working with your own data

Raw files go into `data/raw/`, cleaned ones into `data/processed/`. Always document the source in `data/README.md`.

```{r}
#| eval: false
raw <- read_csv(here("data", "raw", "your_edgelist.csv"))
# ... cleaning steps ...
write_csv(clean, here("data", "processed", "edges_clean.csv"))
```

## Your turn

::: {.callout-tip title="Exercise"}
1. Repeat the cleaning above, but treat the relation as **directed**. How many ties do you get now?
2. Write a short *cleaning log* (3–5 sentences) that justifies every decision you made.
3. Using tidygraph, add a node column that marks isolates (`node_is_isolated()`) and count them per year.
:::

```{r}
# your code here
```
