---
title: "Session 6: Inferential network models"
subtitle: "Exponential random graph models (ERGMs)"
date: today
---

## Goals

- understand why ties cannot be modelled as independent observations,
- read an ERGM as a model of *which local configurations* make a network likely,
- fit, interpret and check ERGMs in R.

::: {.callout-important}
ERGMs live in the **statnet** family of packages, which use their own `network` objects and clash with `igraph` function names. Start this session in a fresh R session (Session → Restart R) and do **not** load igraph. To convert between the two formats, use `intergraph::asNetwork()` and `intergraph::asIgraph()`.
:::

```{r}
library(ergm)
data(florentine)   # loads flomarriage and flobusiness
flomarriage
plot(flomarriage, displaylabels = TRUE)
```

Marriage ties among Renaissance Florentine families: a classic network of alliance politics.

## 1. From the random graph to a logit

The simplest ERGM contains only an `edges` term. It is the random graph from Session 5, written as a statistical model.

```{r}
m1 <- ergm(flomarriage ~ edges)
summary(m1)
```

::: {.callout-note title="Think first"}
Convert the `edges` coefficient into a probability with `plogis()`. Compare it to the network density. Why are they the same?
:::

```{r}
plogis(coef(m1))
network.density(flomarriage)
```

## 2. Adding node attributes

Do wealthier families have more marriage ties?

```{r}
m2 <- ergm(flomarriage ~ edges + nodecov("wealth"))
summary(m2)
```

So far, ties are still treated as independent given the attributes. This model is a logistic regression on all pairs of families.

## 3. Adding dependence: triangles

Now the key step: does a tie become more likely if two families already share a partner? The `gwesp` term captures this tendency towards triangle closure.

```{r}
m3 <- ergm(
  flomarriage ~ edges + nodecov("wealth") + gwesp(decay = 0.25, fixed = TRUE),
  control = control.ergm(seed = 2027)
)
summary(m3)
```

::: {.callout-warning title="Discuss"}
Ties now depend on other ties. That is why ERGMs are estimated by simulation (MCMC) rather than by plain maximum likelihood. Connect this to the dependence problem from Session 1.
:::

## 4. Does the model reproduce the network?

Goodness of fit: simulate networks from the fitted model and compare their statistics with the observed network.

```{r}
fit <- gof(m3)
plot(fit)
```

If the observed values (black line) lie within the simulated range (boxplots), the model captures that feature.

## Your turn

::: {.callout-tip title="Exercise"}
1. Fit the same three models to `flobusiness` (business ties). Do the conclusions change?
2. Add `nodematch()` for a categorical attribute in your own data to test homophily.
3. In two sentences each, interpret the `wealth` and `gwesp` coefficients of your final model for a political science audience.
:::

```{r}
# your code here
```
